\section{Model Exercise 1-1 (05): Swelling of clay}
\label{sec:mex05}
%------------------------------------------------------------------------------
\Authors{Amir Sattari, Keita Yoshioka, Mattias Nest}
%------------------------------------------------------------------------------
Model Exercise (MEX 1-1) investigates the swelling characteristics of the sandy Opalinus claystone. The numerical and experimental approaches are considered to understand the governing factors, which result in materials behavior change, during the swelling process. 


\subsection{Experimental set-up}

\todo[inline]{[IfG](MN) small description of the experimental procedure and the results}

%------------------------------------------------------------------------------


%------------------------------------------------------------------------------
\subsection{Model approach}
\subsubsection*{Lattice element model}

With the application of the lattice model, the simulation of the swelling process in Opalinus Claystone sample is investigated. The experimental data provided by IfG Leipzig are used to determine the heave, swelling pressure and the change of permeability. The expansion of the elements based on the shrinkage and swelling model described in section \ref{Section:ShrinkageLattice} is carried out. The expansion of the elements results in decrease of the hydraulic aperture and therefore lower permeability values. However, during the swelling process the micro fracking is also observed. The elements expansion lead to higher axial confinement stresses between the Voronoi cells, which is represented by interface elements. The linear strains parallel and perpendicular to the embedded layers found in section \ref{sec:mex06} are used to determine the swelling coefficient values. Eventually, the outcome of the simulation is compared to the experimental results. 

\todo[inline]{[CAU] Please add LEM results}

\todo[inline]{Amir: Keita Will you contribute to this MEX?}

%------------------------------------------------------------------------------
\subsection{Results and discussion}